---
title: "bcEpiRate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bcEpiRate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 🔎 Purpose

In population health surveillance and epidemiology, we often want to
measure events of interest (e.g., new cases of a disease, doctor's
visits, hospitalizations) and express them as rates, i.e., the number of
occurrences in a population of a given size. We also often want to
compare these rates across different populations, which may vary in size
and/or demographic composition. These comparisons can be made using
standardization. The `bcEpiRate` package is designed to help users with
these calculations. The aim of this vignette is to demonstrate how to
use the different functions that make up this package.

## 📦 Load packages

Let's begin by loading the necessary packages:

```{r setup, message = FALSE}
library(bcEpiRate)
library(dplyr)
```

This package plays nicely with [`dplyr`](https://dplyr.tidyverse.org/)
functions. In particular, we will be using `group_by()`, `mutate()`, and
`summarise()` frequently in this vignette.

## 📁 Built-in data sets

We will use the following built-in data sets to explore this package:
`counts` , `population`, `std_population`, and `death_rates`.

`counts` contains *simulated* number of deaths by year, cause of death,
and demographic characteristics (i.e., age group and sex).

```{r show-counts}
head(counts)
```

`population` contains the number of people in the *simulated* study
population by year and demographic characteristics.

```{r show-population}
head(population)
```

`std_population` contains the number of people in the *simulated*
standard population by demographic characteristics. These numbers act as
weights that can be used to calculate age- and sex-adjusted statistics.

```{r show-std-population}
head(std_population)
```

The final data set, `death_rates` contains *simulated* age-specific
mortality rates for different causes of death.

```{r show-death-rates}
head(death_rates)
```

In practice, we often source these data sets from different sources
(e.g., Statistics Canada, BC Stats, ministry data warehouse).

We are now ready to start calculating rates and ratios.

## Example 1: Calculating age-specific mortality rates for Disease A in 2018

The **age-specific mortality rate** is calculated by dividing the number
of deaths in a particular age group by the number of people in that age
group in the population (1).

Let's calculate age-specific mortality rates for Disease A in 2018.
First, we will wrangle the built-in data frames:

```{r example-01-wrangle}
# filter number of deaths by disease and year of interest
counts_01 <- counts %>%
  filter(disease == "Disease A", year == 2018) %>%
  group_by(age_group) %>% # aggregate number of deaths by age group
  summarise(counts = sum(counts), .groups = "drop")

# filter study population size by year of interest
popn_01 <- population %>%
  filter(year == 2018) %>%
  group_by(age_group) %>% # aggregate population count by age group
  summarise(popn = sum(popn), .groups = "drop")
 
# join data frames on age group
df_01 <- counts_01 %>%
  left_join(popn_01, by = "age_group")
head(df_01)
```

We will use `get_spec_rt()` to add the age-specific mortality rates (per
100,000 population) as a new column in the data frame:

```{r example-01-get-spec-rt}
df_01 %>%
  mutate(age_spec_rt_per_100k = get_spec_rt(counts = counts, popn = popn, scale = 100000)) %>%
  head()
```

Arguments can also be provided separately as vectors:

```{r example-01-get-spec-rt-02}
age_spec_rt_per_100k <- get_spec_rt(counts = df_01$counts, popn = df_01$popn, scale = 100000)
age_spec_rt_per_100k
```

**Note**: `get_spec_rt()` assumes that the values of the input arguments
`counts` and `popn` are aligned. For example, if the number of deaths
observed in the 0-4 age group appears first in the vector passed into
`counts`, the number of people in this age group must appear first in
`popn`. This logic applies to the other functions in this package.

## Example 2: Calculating sex-specific mortality rates for Disease A in 2018

We can also calculate mortality rates for a specific sex.

Let's calculate sex-specific mortality rates for Disease A in 2018.
Again, we will first wrangle the data frames:

```{r example-02-wrangle}
# filter number of deaths by disease and year of interest
counts_02 <- counts %>%
  filter(disease == "Disease A", year == 2018) %>%
  group_by(sex) %>% # aggregate number of deaths by sex
  summarise(counts = sum(counts), .groups = "drop")

# filter population size by year of interest
popn_02 <- population %>%
  filter(year == 2018) %>%
  group_by(sex) %>% # aggregate population count by sex
  summarise(popn = sum(popn), .groups = "drop")

# join data frames on sex
df_02 <- counts_02 %>%
  left_join(popn_02, by = "sex")
df_02
```

Use `get_spec_rt()` to calculate the sex-specific mortality rates (per
100,000 population) and construct 95% confidence intervals based on the
Normal distribution.

```{r example-02-get-spec-rt, message = FALSE}
df_02 <- df_02 %>%
  mutate(get_spec_rt(counts = counts, popn = popn, scale = 100000, dist = "normal", interval = 0.95))
df_02
```

```{r example-02-text, include=FALSE}
# store all values for summary
df_02 <- df_02 %>%
  mutate(across(.cols = c(counts, popn), ~ format(.x, big.mark = ",")),
         across(.cols = c(rate, lower, upper), ~ round(.x, 1)))
df_02_f <- df_02 %>% filter(sex == "F")
df_02_m <- df_02 %>% filter(sex == "M")
```

The `rate` column contains the estimate of the specific rate and the
`lower` and `upper` columns contain the confidence limits. For females,
`r df_02_f$counts` deaths were observed in a population of
`r df_02_f$popn`. This corresponds to a sex-specific mortality rate of
`r df_02_f$rate` per 100,000 population; 95% CI [`r df_02_f$lower`,
`r df_02_f$upper`]. In comparison, for males, `r df_02_m$counts` deaths
were observed in a population of `r df_02_m$popn`. This corresponds to a
rate of `r df_02_m$rate` per 100,000 population; 95% CI
[`r df_02_m$lower`, `r df_02_m$upper`].

## Example 3: Calculating the crude mortality rate for Disease B in 2019

The **crude mortality rate** is calculated by dividing the total number
of deaths by the total number of people in a population (2).

In addition to specific rates, `get_spec_rt()` can be used to calculate
crude rates, provided `counts` and `popn` are scalars (i.e., single
integer values) which represent the total number of deaths and the size
of the population respectively.

Let's calculate the crude mortality rate (per 100,000) for Disease B in
2019:

```{r example-03}
# total number of deaths caused by Disease B in 2019
counts_03 <- counts %>%
  filter(disease == "Disease B", year == 2019) %>%
  summarise(counts = sum(counts)) %>%
  pull(counts) 

# size of the population in 2019
popn_03 <- population %>%
  filter(year == 2019) %>%
  summarise(popn = sum(popn)) %>%
  pull(popn)

df_03 <- data.frame(counts = counts_03, popn = popn_03) %>%
  mutate(get_spec_rt(counts = counts, popn = popn, scale = 100000, dist = "normal", interval = 0.95))
df_03
```

```{r example-03-text, include=FALSE}
# store all values for summary
df_03 <- df_03 %>%
  mutate(across(.cols = c(counts, popn), ~ format(.x, big.mark = ",")),
         across(.cols = c(rate, lower, upper), ~ round(.x, 1)))

```

In 2019, `r df_03$counts` deaths from Disease B were reported in a
population of `r df_03$popn`. This corresponds to a crude mortality rate
of `r df_03$rate` per 100,000 population; 95% CI [`r df_03$lower`,
`r df_03$upper`].

## Example 4: Calculating the age-standardized mortality rate for Disease C in 2019

The **age-standardized mortality rate** is a weighted average of the
age-specific mortality rates, where the weights are derived from the age
structure of the standard population (3). This procedure is also
referred to as **direct standardization**.

Statistics Canada provides an example as to why it is important to use
standardized rates when comparing different populations (4):

> [The] 2011 Canadian population has a higher proportion of those 40+
> than the 2000 population does: almost half (49.9%) of the 2011
> population was 40 years of age or older, compared to 44.4% in 2000.
> Due to the high mortality rate in the 40+ age group, considerably more
> cancer deaths are observed in 2011. But, it is only by removing the
> effect of the differing age distributions that we can make conclusions
> about the relative decreases or increases in mortality over time.

Let's calculate the age-standardized mortality rate for Disease C in
2019. We will first wrangle `counts`, `population`, and
`std_population`:

```{r example-04-wrangle}
# filter number of deaths by disease and year of interest
counts_04 <- counts %>%
  filter(disease == "Disease C", year == 2019) %>%
  group_by(age_group) %>% # aggregate number of deaths by age group
  summarise(counts = sum(counts), .groups = "drop")
  
# filter *study* population size by year of interest
popn_04 <- population %>%
  filter(year == 2019) %>%
  group_by(age_group) %>% # aggregate population count by age group
  summarise(popn = sum(popn), .groups = "drop")

# wrangle *standard* population
std_popn_04 <- std_population %>%
  group_by(age_group) %>% # aggregate population count by age group
  summarise(std_popn = sum(std_popn), .groups = "drop")

# join data frames on age group
df_04 <- counts_04 %>%
  left_join(popn_04, by = "age_group") %>%
  left_join(std_popn_04, by = "age_group")
head(df_04)
```

Each row contains the number of deaths from Disease C, the number of
people in the study population, and the number of people in the standard
population for a particular age group.

Next, we can use `get_ds_rt()` to calculate the age-standardized
mortality rate. We will also construct a 95% confidence interval around
the estimate:

```{r example-04-get-ds-rt}
df_04 <- df_04 %>%
  summarise(get_ds_rt(
    counts = counts, popn = popn, std_popn = std_popn,
    scale = 100000, dist = "normal", interval = 0.95
  ))
df_04
```

```{r example-04-text, include=FALSE}
# store all values for summary
df_04 <- df_04 %>%
  mutate(across(.cols = c(dsr, lower, upper), ~ round(.x, 1)))
df_04
```

The age-standardized mortality rate for Disease C in 2019 was
`r df_04$dsr` per 100,000 population; 95% CI [`r df_04$lower`,
`r df_04$upper`].

## Example 5: Calculating multiple age-standardized mortality rates at once

We often want to calculate age-standardized mortality rates for multiple
diseases over the years. We will explore how to do this using
`get_ds_rt()` and `dplyr` functions.

First, let's wrangle `counts`, `population`, and `std_population`:

```{r example-05-wrangle}
# aggregate number of deaths by year, cause of death, and age group
counts_05 <- counts %>%
  group_by(year, disease, age_group) %>%
  summarise(counts = sum(counts), .groups = "drop")

# aggregate *study* population size by year and age group
popn_05 <- population %>%
  group_by(year, age_group) %>%
  summarise(popn = sum(popn), .groups = "drop")

# aggregate *standard* population size by age group
std_popn_05 <- std_population %>%
  group_by(age_group) %>%
  summarise(std_popn = sum(std_popn), .groups = "drop")

# join data frames
df_05 <- counts_05 %>%
  left_join(popn_05, by = c("year", "age_group")) %>%
  left_join(std_popn_05, by = "age_group")
head(df_05)
```

To calculate age-standardized mortality rates for all combinations of
`year` and `disease`, we call `group_by()` followed by `summarise()`
along with `get_ds_rt()`. Let's construct 95% confidence intervals
around each estimate as well:

```{r example-05-get-ds-rt}
df_05 <- df_05 %>%
  group_by(year, disease) %>%
  summarise(get_ds_rt(
    counts = counts, popn = popn, std_popn = std_popn,
    scale = 100000, dist = "normal", interval = 0.95
  ), .groups = "drop")
df_05
```

In just a few lines of code, we have calculated age-standardized
mortality rates (per 100,000 population) along with their confidence
intervals for 5 different causes of death over 5 years.

## Example 6: Calculating the standard mortality ratio for Disease D in 2016

The **standard mortality ratio (SMR)** is a ratio between the number of
observed deaths in the study population and the number of expected
deaths in the standard population. When the ratio is greater than 1, it
suggests that there is excess mortality in the study population. On the
other hand, when it is less than 1, it suggests that the number of
deaths in the study population is fewer than what was expected (5).

To calculate the SMR for Disease D in 2016, we will need the number of
observed deaths, the size of each age group in the study population, and
the age-specific mortality rate for each subgroup in the standard
population. Let's say that 278 people died from Disease D in 2016:

```{r example-06}
# define number of observed deaths


# filter study population size by year of interest

# filter age-specific mortality rates by disease of interest

  
# calculate SMR

```

## 🤹 Exception handling

`get_spec_rt()`, `get_ds_rt()`, and `get_smr()` are designed to handle
edge cases. Access the documentation pages for more information by using
either `help()` or `?`.

## 📚 References

1.  [Principles of Epidemiology \| Lesson 3 - Section
    3](https://www.cdc.gov/csels/dsepd/ss1978/lesson3/section3.html)
2.  [Crude and Age-Adjusted Rate \| Health & Senior
    Services](https://health.mo.gov/data/documentation/crude-aarate.php)
3.  [Age-standardized mortality rate (per 100 000
    population)](https://www.who.int/data/gho/indicator-metadata-registry/imr-details/78)
4.  [Age-standardized Rates](https://www.statcan.gc.ca/eng/dai/btd/asr)
5.  [Standardized mortality ratio -
    Wikipedia](https://en.wikipedia.org/wiki/Standardized_mortality_ratio)
